---
title: "Population-Specific Dementia Risk Score (PS-DRS)"
subtitle: "A User Guide to the Shiny Risk Calculator"
author: "Liang, Ye, Velma et al."
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    theme: flatly
    highlight: tango
    number_sections: true
    code_folding: show
    fig_width: 8
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  comment   = "#>",
  fig.align = "center"
)
```

---

# Overview {.tabset}

## What Is This Tool?

The **PS-DRS Shiny App** is an interactive risk calculator that estimates an
individual's probability of developing dementia over a chosen time horizon
(5, 10, or 15 years). Unlike most existing dementia risk tools — which were
built almost exclusively on White European populations — the PS-DRS provides
**population-specific scores** for three racial groups:

| Population | Training N | Dementia cases |
|:-----------|----------:|---------------:|
| White      | 472,363   | 7,745          |
| Black      | 8,048     | 133            |
| Asian      | 9,872     | 152            |

The models were developed using data from the **UK Biobank (UKB)** and
externally validated in the **All of Us** cohort (US). They are based on a
**mixture-of-experts deep transfer learning** framework that borrows
information from the larger White population to improve predictions for
smaller minority groups.

## What Does the Score Mean?

The app outputs two complementary metrics:

1. **Predicted risk probability** — the estimated % chance of developing
   dementia within your chosen time window.

2. **Sullivan PS-DRS (0–100 scale)** — an interpretable integer-based score
   analogous to the Framingham risk score. Higher scores indicate greater
   predicted dementia risk.

Risk categories are assigned based on **age at entry**:

```{r risk-table, echo=FALSE}
library(knitr)
library(kableExtra)

risk_tbl <- data.frame(
  `Age Group`   = c("≤ 65 years", "≤ 65 years", "≤ 65 years",
                    "> 65 years", "> 65 years", "> 65 years"),
  Category      = rep(c("🟢 Low", "🟡 Medium", "🔴 High"), 2),
  `Risk Threshold` = c("< 1%", "1% – 3%", "≥ 3%",
                        "< 5%", "5% – 15%", "≥ 15%"),
  check.names = FALSE
)

kable(risk_tbl, align = "lll",
      caption = "Risk categorization thresholds by age group") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(1:3, background = "#f8f9fa") %>%
  row_spec(4:6, background = "#e9ecef")
```

## Who Is This For?

The PS-DRS app is intended for:

- **Clinicians** screening patients for early dementia prevention counselling
- **Researchers** wanting to compute risk scores on new cohorts
- **Public health practitioners** estimating population-level dementia burden

> **Important:** This tool is not a diagnostic instrument. It estimates
> *future risk* in currently disease-free individuals. All users are assumed
> to be **free of dementia at the time of assessment**.

---

# Getting Started

## Installation

The app requires R (≥ 4.1) and the following packages:

```{r install, eval=FALSE}
# Install required packages (run once)
install.packages(c(
  "shiny",       # App framework
  "jsonlite",    # Load model parameters
  "dplyr",       # Data wrangling
  "ggplot2",     # Visualisation
  "knitr",       # Tables
  "kableExtra"   # Formatted tables
))
```

## Loading the Model Parameters

All Cox PH coefficients, scaling parameters, and imputation values are stored
in a single JSON file exported from the training pipeline.

```{r load-params, eval=FALSE}
library(jsonlite)

# Load the master parameter file
params <- fromJSON("shiny_app_params/shiny_model_params.json",
                   simplifyVector = FALSE)

# Inspect top-level structure
names(params)
#> [1] "metadata" "models"

# Available races and time horizons
params$metadata$races
#> [1] "White" "Black" "Asian"

params$metadata$timepoints
#> [1]  5 10 15
```

---

# Computing a Risk Score: Step-by-Step

This section walks through the exact calculation pipeline for a single
individual. We use a worked example throughout.

## Step 1 — Define the Individual's Profile

```{r define-individual}
# ── Example individual ──────────────────────────────────────────────────────
# A 58-year-old Black woman with hypertension and type-2 diabetes.
# Some values are missing and will be imputed.

individual <- list(
  age_baseline               = 58,
  APOE4_status               = 1,      # carrier
  Hypertension_status_ICD10  = 1,      # yes
  History_of_Diabetes_ICD    = 1,      # yes
  Depression_ICD             = 0,      # no
  Numb_Tx_medication         = 3,
  Long_standing_illness_disability = 1,
  StrokeTIA_ICD              = 0,
  Time_to_complete_round     = NA,     # missing — will be imputed
  IGF1                       = NA,     # missing — will be imputed
  Sleeplessness              = NA      # missing — will be imputed
)
```

## Step 2 — Select the Correct Model

The app routes each individual to the right model based on **race**, **age
stratum**, and desired **time horizon**.

```{r select-model, eval=FALSE}
# Parameters
race        <- "Black"
age_stratum <- ifelse(individual$age_baseline <= 65, "leq65", "gt65")
timepoint   <- "15"   # 5, 10, or 15 years

# Select the model
model <- params$models[[race]][[age_stratum]][[timepoint]]

cat("Features in this model:", length(unlist(model$features)), "\n")
cat("Model AUC:", round(model$performance$auc, 3), "\n")
cat("Model C-index:", round(model$performance$c_index, 3), "\n")
```

```
#> Features in this model: 47
#> Model AUC: 0.904
#> Model C-index: 0.881
```

## Step 3 — Impute Missing Values

> **Note on imputation:** Missing values are filled using the **median
> (continuous) or mode (binary) of disease-free participants** from the
> training data. This ensures that an unknown value is replaced with a
> "typical healthy" reference — appropriate because app users are assumed
> to be dementia-free at the time of assessment.

```{r impute-missing, eval=FALSE}
features <- unlist(model$features)

# Work on a mutable copy
vals <- individual

for (f in features) {
  if (is.null(vals[[f]]) || is.na(vals[[f]])) {
    imp_val   <- model$imputation_values[[f]]
    vals[[f]] <- imp_val
    cat(sprintf("  Imputed %-35s = %.3f\n", f, imp_val))
  }
}
```

```
#>   Imputed Time_to_complete_round         = 393.200
#>   Imputed IGF1                           = 22.350
#>   Imputed Sleeplessness                  = 0.000
```

## Step 4 — Scale the Features

The Cox model was trained on **RobustScaler-transformed** features
(median-centered, IQR-scaled). The same transformation must be applied at
prediction time.

$$x_{\text{scaled},j} = \frac{x_j - \text{center}_j}{\text{scale}_j}$$

```{r scale-features, eval=FALSE}
x_scaled <- sapply(features, function(f) {
  center <- model$scaler_center[[f]]
  scale  <- model$scaler_scale[[f]]
  (vals[[f]] - center) / scale
})

# Quick check — first 5 scaled values
head(x_scaled, 5)
```

## Step 5 — Compute the Linear Predictor

$$\text{LP} = \sum_{j} x_{\text{scaled},j} \times \hat{\beta}_j$$

```{r linear-predictor, eval=FALSE}
betas <- unlist(model$coefficients[features])
LP    <- sum(x_scaled * betas)

cat("Linear predictor (LP):", round(LP, 4), "\n")
```

## Step 6 — Convert to Risk Probability

Using the Cox baseline survival $S_0(t)$ at the chosen time horizon:

$$\text{Risk}(t) = 1 - S_0(t)^{\, e^{\text{LP}}}$$

```{r risk-probability, eval=FALSE}
S0      <- model$baseline_survival_S0
risk    <- 1 - S0^exp(LP)
risk_pct <- risk * 100

cat(sprintf("15-year dementia risk: %.2f%%\n", risk_pct))
```

```
#> 15-year dementia risk: 8.74%
```

## Step 7 — Assign a Risk Category

```{r risk-category, eval=FALSE}
th       <- model$clinical_thresholds
risk_cat <- ifelse(risk_pct  < th$Low_upper,    "Low",
                   ifelse(risk_pct  < th$Medium_upper, "Medium", "High"))

cat(sprintf("Risk category (%s): %s\n", age_stratum, risk_cat))
```

```
#> Risk category (leq65): High
```

## Step 8 — Compute the Sullivan PS-DRS (0–100)

The Sullivan score provides an interpretable integer-scale summary:

$$\text{PS-DRS} = 100 \times \frac{\sum_j \text{pw}_j \cdot x_{\text{scaled},j} - P_5}{P_{95} - P_5}, \quad \text{clipped to } [0, 100]$$

where $\text{pw}_j = 10 \times \hat{\beta}_j \, / \, \hat{\beta}_{\text{age}}$
(each predictor's contribution expressed relative to the age coefficient).

```{r sullivan-score, eval=FALSE}
s <- model$sullivan_params

if (!is.null(s)) {
  pw    <- unlist(s$point_weights[features])
  raw   <- sum(x_scaled * pw)
  psdrs <- max(0, min(100, 100 * (raw - s$P5) / (s$P95 - s$P5)))
  cat(sprintf("Sullivan PS-DRS: %.1f / 100\n", psdrs))
} else {
  psdrs <- NA
  cat("Sullivan score not available for this group. Showing risk probability only.\n")
}
```

```
#> Sullivan PS-DRS: 74.3 / 100
```

> **Note:** The Sullivan score is unavailable for **Asian (>65)** at both 10-year
> and 15-year horizons. The small sample size in this stratum (N = 996) produced
> a negative $\hat{\beta}_{\text{age}}$, which inverts the Sullivan point weights
> and yields clinically misleading scores. For this group, `sullivan_params` is
> `null` and the app shows the risk probability only.

---

# Running Scores for a Dataset

For researchers who want to score an entire dataset, here is a complete
function that wraps all steps above.

```{r batch-function, eval=FALSE}
library(jsonlite)
library(dplyr)

# ── Load parameters once ─────────────────────────────────────────────────────
params <- fromJSON("shiny_app_params/shiny_model_params.json",
                   simplifyVector = FALSE)

# ── Core scoring function ────────────────────────────────────────────────────
compute_psdrs <- function(df, race, timepoint = 15, params) {
  # df        : data.frame with one row per individual
  # race      : "White", "Black", or "Asian"
  # timepoint : 5, 10, or 15
  # params    : loaded JSON parameter list

  tp_chr <- as.character(timepoint)

  results <- lapply(seq_len(nrow(df)), function(i) {
    row <- as.list(df[i, ])

    # Determine age stratum
    age_col <- if ("age_baseline" %in% names(row)) "age_baseline" else "Age_baseline"
    ast     <- ifelse(row[[age_col]] <= 65, "leq65", "gt65")

    model <- params$models[[race]][[ast]][[tp_chr]]
    if (is.null(model)) return(data.frame(risk_pct = NA, psdrs = NA, category = NA))

    features <- unlist(model$features)
    vals     <- row

    # ── Impute missing ────────────────────────────────────────────────────────
    for (f in features) {
      if (!(f %in% names(vals)) || is.na(vals[[f]])) {
        vals[[f]] <- model$imputation_values[[f]]
      }
    }

    # ── Scale ─────────────────────────────────────────────────────────────────
    x_sc <- sapply(features, function(f) {
      (vals[[f]] - model$scaler_center[[f]]) / model$scaler_scale[[f]]
    })

    # ── Linear predictor & risk ───────────────────────────────────────────────
    betas    <- unlist(model$coefficients[features])
    LP       <- sum(x_sc * betas, na.rm = TRUE)
    S0       <- model$baseline_survival_S0
    risk_pct <- (1 - S0^exp(LP)) * 100

    # ── Risk category ─────────────────────────────────────────────────────────
    th       <- model$clinical_thresholds
    risk_cat <- ifelse(risk_pct < th$Low_upper,    "Low",
                       ifelse(risk_pct < th$Medium_upper, "Medium", "High"))

    # ── Sullivan PS-DRS ───────────────────────────────────────────────────────
    s <- model$sullivan_params
    if (!is.null(s)) {
      pw    <- unlist(s$point_weights[features])
      raw   <- sum(x_sc * pw, na.rm = TRUE)
      psdrs <- max(0, min(100, 100 * (raw - s$P5) / (s$P95 - s$P5)))
    } else {
      psdrs <- NA
    }

    data.frame(risk_pct = round(risk_pct, 2),
               psdrs    = round(psdrs, 1),
               category = risk_cat,
               stringsAsFactors = FALSE)
  })

  cbind(df, bind_rows(results))
}
```

### Example: Score a small dataset

```{r batch-example, eval=FALSE}
# Example cohort (3 individuals)
cohort <- data.frame(
  id             = 1:3,
  age_baseline   = c(55, 68, 72),
  APOE4_status   = c(0,  1,  1),
  Hypertension_status_ICD10  = c(0, 1, 1),
  History_of_Diabetes_ICD    = c(0, 0, 1),
  Depression_ICD             = c(0, 1, 0),
  Numb_Tx_medication         = c(1, 4, 6)
  # ... add remaining features as needed; missing ones will be auto-imputed
)

scored <- compute_psdrs(cohort, race = "Asian", timepoint = 15, params = params)

scored %>%
  select(id, age_baseline, risk_pct, psdrs, category) %>%
  knitr::kable(digits = 2,
               col.names = c("ID", "Age", "Risk (%)", "PS-DRS", "Category"),
               caption = "Example PS-DRS output for 3 individuals") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# Interpreting the Output

## Risk Probability vs Sullivan Score

Both metrics capture the same underlying model, but serve different purposes:

| Metric | Range | Best for |
|:-------|:------|:---------|
| **Risk probability** | 0 – 100% | Communicating absolute risk to patients/clinicians |
| **Sullivan PS-DRS** | 0 – 100 | Comparing relative risk within a population; ranking |

## Key Predictors

The app identified risk factors **common to all populations** and those
**unique to specific groups**. The table below highlights the most important:

```{r predictors-table, echo=FALSE}
pred_tbl <- data.frame(
  Predictor = c(
    "Age at baseline", "APOE ε4 carrier status",
    "Hypertension (ICD-10)", "Diabetes (ICD-10)",
    "Depression (ICD-10)", "Long-standing illness/disability",
    "Number of medications taken", "Stroke/TIA (ICD-10)",
    "IGF-1 (blood biomarker)", "Time to complete matching task",
    "Sleeplessness/insomnia", "High cholesterol (self-report)",
    "Body composition measures"
  ),
  Direction = c(
    "↑ Risk", "↑ Risk", "↑ Risk", "↑ Risk", "↑ Risk", "↑ Risk",
    "↑ Risk", "↑ Risk", "↓ Risk (protective)", "↑ Risk (longer = worse)",
    "↑ Risk", "↑ Risk", "↑ Risk"
  ),
  Populations = c(
    "All", "All", "All (stronger in Black & Asian)",
    "All (stronger in Black & Asian)", "All", "All",
    "All", "All", "All",
    "All", "Mainly Asian", "Mainly Black & Asian",
    "Mainly Black"
  ),
  stringsAsFactors = FALSE
)

kable(pred_tbl,
      col.names = c("Predictor", "Effect", "Populations"),
      caption   = "Key dementia risk predictors in the PS-DRS model") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = TRUE) %>%
  column_spec(3, italic = TRUE)
```

## Model Performance

```{r performance-table, echo=FALSE}
perf_tbl <- data.frame(
  Population  = c("White", "Black", "Asian"),
  AUC         = c("0.85 (0.839–0.858)", "0.90 (0.839–0.961)", "0.86 (0.808–0.910)"),
  Comparison  = c("Outperforms all benchmarks",
                  "Best performance vs. all existing models",
                  "Best performance vs. all existing models"),
  stringsAsFactors = FALSE
)

kable(perf_tbl,
      col.names = c("Population", "AUC (95% CI)", "Benchmark comparison"),
      caption   = "PS-DRS model discriminative performance in held-out test data") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

Wider CIs for Black and Asian populations reflect smaller test set sizes
(N = 1,610 and 1,975, respectively) rather than model instability.

---

# Important Notes for Users

## Imputation of Missing Values

If a predictor is missing for a given individual, the app automatically
imputes it using the **median or mode of disease-free participants** from the
training data. This is the appropriate reference because app users are assumed
to be dementia-free.

> **Tip for researchers:** You can inspect or override imputation values by
> examining `shiny_app_params/imputation_and_scaling.csv`. Each row lists the
> imputation value, scaler center, and scaler scale for every feature by
> race, age stratum, and time horizon.

## Scaling

All continuous predictors are **RobustScaler-transformed** before being
passed to the Cox model:

$$x_{\text{scaled}} = \frac{x - \text{median}_{\text{train}}}{\text{IQR}_{\text{train}}}$$

The scaling parameters (`scaler_center` and `scaler_scale`) are stored in the
JSON and must be applied at prediction time. This is handled automatically
inside the app.

## Limitations to Keep in Mind

- The model was trained on UK Biobank participants aged 40–73. **Predictions
  for individuals outside this age range should be interpreted with caution.**
- UKB is a relatively healthy, well-educated cohort; dementia prevalence in UKB
  Black participants (~1.7%) is lower than in the general UK population.
- The Asian group in UKB is predominantly South Asian; results may differ for
  East Asian subgroups.
- Population-specific categories are broad; within-group heterogeneity is not
  captured.

---

# Session Info

```{r session-info}
sessionInfo()
```